Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2019-01-31 accessPEP.Rmd
updated docs for 1/30 meeting with Jess and Katie

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
---

@@

## To do
- Add in comparison to district model (+ demonstrate why estimating burden at district level does not result in realistic estimates of burden...because reporting for Moramanga District will be estimated at 0.98, same as for Antananarivo Renivohitra (i.e. Tana)
- Sensitivity analyses for model: comparison to null models, with variation in reporting estimate
- Sensitivity analyses for burden: comparison to null models (flat reporting, ttimes at district level) p_rabid (places with high ttimes), rho_max (places with high ttimes), R (scaling with pop); 
- Calculate rho_max from Moramanga data
- Calculate p_rabid from Moramanga data (not to include in model but as supplementary)

## INTRODUCTION
Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis. However, access to this intervention is highly limited in areas where the disease is endemic, often due to inaccessibility to health centres which provide PEP (*cite GAVI paper or Nandini's paper of geographic availability of PEP*). Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.

@@

The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space. While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimates rabies deaths in places with high reporting of bites. The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level. Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP and the endemic circulation of canine rabies, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.

@@

### GIS Data
We used the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the commune level.

@@

## Centers with no data ## Change this to be so that only ones with zero forms!
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Marolambo", "Nosy be", "Sainte Marie", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

@@

### Bite patient data
We used a database of bite patient forms submitted to IPM from 24 ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted as frequently as monthly to annually by clinics. These data include the administrative district of the bite patient's address and the date of reporting.

As low risk contacts with probable or confirmed rabies cases have been shown to make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018), we attempted to exclude these by looking at the distribution of patients reporting per day to a given clinic. Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B). For the national data for which a subset of patient forms were noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts. We further excluded known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data.

ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
  geom_jitter(alpha = 0.5, width = 0.25) +
  coord_flip()

plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
points(check$sd, check$prop_bites, col = "red", pch = 20)
```

@@

@@

## To plot lines between ctar + districts
exp_mat %>% 
  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.2
```

```{r figure 1A}
pdf("figs/network.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, 
         y1 = dist_lines$LATITUDE,
         lwd = dist_lines$size*2, col = dist_lines$color)
points(ctar_points$LONGITUDE, ctar_points$LATITUDE, col = ctar_points$color, pch = 1,
       cex = log(ctar_points$n + 0.1)*0.25)
points(dist_points$dist_long, dist_points$dist_lat, cex = dist_points$size, pch = 20,
       col = dist_points$color)
dev.off()
```

```{r network output, echo=FALSE, fig.cap="Figure 1C", results = "show", out.width = '100%'}
knitr::include_graphics("figs/network.pdf")
```

@@

For each clinic, we defined the catchment area as all districts for which the clinic was the closest ARMC. For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A).

@@

```{r figure 2B, fig.cap = "Figure 2B"}

clinic_reporting %>%
  mutate(prop15 = prop, prop5 = get.days(threshold = 5)[[1]]$prop,
         prop30 = get.days(threshold = 30)[[1]]$prop) %>%
  ggplot(., aes(x = reorder(ctar, prop), color = ctar)) +
  geom_hline(yintercept = 0.25, color = "darkgrey") +
  geom_boxplot(aes(ymin = prop30, lower = prop15, middle = prop15, upper = prop15, ymax = prop5),
               stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  xlab("") +
  ylab("Proportion of days included") +
  coord_flip()

@@

As there were substantial undersubmission of forms from clinics *...* need to describe the reason why we think this is temporal, we excluded any periods of time for which there were no forms submitted for 15 consecutive days. We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure 2A). This estimate did vary based on our assumption of the threshold number of consecutive days (Figure 2B), so we looked at sensitivity of parameter estimates to changing this threshold (Supplementary Materials?). In all cases, we further excluded any years for which there was less than 25% reporting.

@@

@@

Our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).

### Model of reported bites as a function of travel time
While the national bite patient data is available at the district level, travel times and therefore reporting likely vary significantly at the sub-district level. In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level. Incidence at the commune level is then a function of travel times to the closest ARMC. Specifically, we modeled bites as follows:

@@

Finally, we compared these models to a model of bites at the district level as the function of trave times averaged to the district level:
$$ 
  \mu_{d} = exp(\beta_{t}T_d + \beta_0)\times pop_d 
$$
where $pop_d$, $T_d$, and $mu_d$ are respectively human population size, weighted travel times, and mean number of bites in district $d$. As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models, as a test of whether the observed patterns could be predicted by bite incidence scaling with population size. We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.

@@

### Estimating the impact of expanding PEP Access
We use this framework to compare three scenarios of PEP provisioning in Madagascar:

@@

### Strategically expanding PEP access
Given limited resources and functioning of clinics, we developed a framework to look at the incremental benefit of adding ARMC. Starting with the current locations (scenario 1), we added one clinic and recalculated the proportion of people living < x ttime hours away from a clinic for the country. We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding them sequentially. We calculate burden at each step and look at the incremental reduction in burden as each clinic is added.

### Sensitivity analyses for burden estimates and scenario analyses
- R
- Rho_max
- P_rabid

@@

par_df <- data.frame(model = c("Mada", "Mora", "District", "Mada", "Mora", "District"),
                     par = c("B_ttimes", "B_ttimes", "B_ttimes", "intercept", "intercept",
                             "intercept"),
                     values = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist, 
                                B_0_mada, B_0_mora, B_0_dist),
                     upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_ttimes_dist_CI[2],
                                  B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                     lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  B_ttimes_dist_CI[1], 
                                  B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]))
kable(par_df, caption = "Table 1: summary of parameter estimates")
ttimes_plot <- seq(0, 15*60, by = 30)
pop_plot <- 1e5

@@

legend("topright", c("Mada", "Mora", "District"), col = c("darkblue", "red", "purple"), 
       lty = c(1, 1, 1), 
       bty = "n")
```

@@

```{r figure 5B, fig.cap="Figure 5B"}
## Data vs. predicted
comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
                   xlab = "Observed mean bites", ylab = "Model predicted mean bites")
abline(a = 0, b = 1, col = "grey", lty = 2)

mora_bites %>%
      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("red", 0.5))

exps_dist %>%
      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
points(check$bites, check$bites_mu, col = alpha("blue", 0.5), pch = 20)

library(bbmle)
get.likelihood.bb <- function(commune, district, params, B_ttimes, B_0) {
  # To test likelihood function
  commune = comm_data
  district = exps_dist
  # params = c(1e-6, 0.1)
  commune %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
}
m3 <- mle2(get.likelihood.bb, start = list(B_ttimes = 1e-6, B_0 = 0.1))

```{r pop sensitivity analsyis }
## Pop only
lm(exps_dist$bites ~ exps_dist$pop)

get.likelihood.pop <- function(commune, district, B_pop, B_0) {
  # To test likelihood function
  commune = comm_data
  district = exps_dist
  # params = c(1e-6, 0.1)
  commune %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_pop*(pop/1000) + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
}

### Pop parameters
m3 <- mle2(get.likelihood.pop, start = list(B_pop = 1e-6, B_0 = 0.1))

pop_pars <- optim(par = c(1e-6, 0.1), get.likelihood.pop, district = exps_dist,
             commune = comm_data,
              hessian = TRUE)
B_pop_pop <- pop_pars$par[1]
B_0_pop <- pop_pars$par[2]

## Getting standard errors
# this is how you can get standard errors from optim, but inverting the hessian
se <- sqrt(abs(diag(solve(pop_pars$hessian)))); se

#simulate "standard errors"
varcov <- solve(pop_pars$hessian)

ntest = 1000; testpar = pop_pars$par; sim.par = matrix(NA, ntest, 2)
for (k in 1:ntest) {
  sim.par[k, ] = mvrnorm(1, pop_pars$par, varcov)
}

B_pop_pop_CI <- quantile(sim.par[,1],c(0.025,0.975))
B_0_pop_CI <- quantile(sim.par[,2],c(0.025,0.975))

comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_pop_pop*(pop/1000) + B_0_pop)*pop) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, 
                   col = alpha("darkblue", 0.5),
                   xlab = "Observed mean bites", ylab = "Model predicted mean bites")
abline(a = 0, b = 1, col = "grey", lty = 2)
sum_bites %$% cor(dist_bites, sum_bites)

comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% points(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("magenta", 0.5))
abline(a = 0, b = 1, col = "grey", lty = 2)
sum_bites %$% cor(dist_bites, sum_bites)

@@

We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 5A), with reported bite incidence decreasing with travel times to the ARMC. All of the models produced reasonable fits to the data (Figure 5B,...), however, there was some variation in bite incidence which both models were not able to capture.

- compare to null models: flat incidence, incidence solely predicted by pop;
- sensitivity of param estimates to reporting cut-offs (i.e. days excluded)
- sensitivity of param estimates to exclusion of contacts

@@

```{r functions for getting burden}

boots <- function(x, nboot = 1000){
  n = length(x)
  xbar = mean(x)
  tmpdata = sample(x,n*nboot, replace=TRUE)
  bootstrapsample = matrix(tmpdata, nrow=n, ncol=nboot)
  bsmeans = colMeans(bootstrapsample)
  deltastar = bsmeans - xbar
  d = quantile(deltastar, c(0.05, 0.95), na.rm = TRUE)
  ## Get from quantiles
  ci = xbar - c(d[2], d[1])

  ## Sorted CIs
  sorteddeltastar = sort(deltastar)
  # # Look at the sorted results
  # Find the .05 and .95 critical values of deltastar
  d9alt = sorteddeltastar[5]
  d1alt = sorteddeltastar[950]
  # Find and print the 80% confidence interval for the mean
  ciAlt = xbar - c(d1alt,d9alt)
  # cat(’Alternative confidence interval: ’,ciAlt, ’\n’)
  return(ci)
}

get.meanCI <- function(matrix_by_row, nboots = 1000) {
    mean_rows <- apply(matrix_by_row, 1, mean)
    CIs <- apply(matrix_by_row, 1, boots, nboot = nboots)
    lower <- unlist(CIs)[1, ] 
    upper <- unlist(CIs)[2, ]
    summ <- cbind(mean_rows, lower, upper)
    colnames(summ) <- c("mean", "lower", "upper")
    return(summ)
}

get.burden <- function(covars = comm_data, ttimes = comm_data$ttimes_weighted, pop = comm_data$pop, 
                       param_ttimes = B_ttimes_mada,
                       param_intercept = B_0_mada,
                       p_rabid = 0.5, rho_max = 0.9,
                       max_HDR = 25, min_HDR = 5, 
                       dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
                       prob_death = 0.16, nsims = 1000) {

  store_deaths <- store_p_rabid <- matrix(NA, length(ttimes), nsims)

  for (i in 1:nsims) {
    ## for testing
    # covars = comm_data; ttimes = comm_data$ttimes_weighted; pop = comm_data$pop;
    #                    param_ttimes = B_ttimes_mada;
    #                    param_intercept = B_0_mada;
    #                    p_rabid = 0.5; rho_max = 0.9;
    #                    max_HDR = 25; min_HDR = 5;
    #                    dog_rabies_inc = 0.01; human_exp_rate = 0.39;
    #                    prob_death = 0.16;

    bites <- rpois(length(ttimes), exp(param_ttimes*ttimes + param_intercept)*pop)
    mean_exps <- runif(length(pop), min = human_exp_rate*dog_rabies_inc*pop/max_HDR, 
                    max = human_exp_rate*dog_rabies_inc*pop/min_HDR)
    rabid_exps <- rpois(length(mean_exps), mean_exps)
    p_rabid_t <- ifelse(bites == 0 & rabid_exps == 0, 0,
                        ifelse(bites*p_rabid/rabid_exps > rho_max, (rho_max*rabid_exps)/bites, p_rabid))
    reported_rabid <- round(bites*p_rabid_t)
    reported_rabid <- ifelse(reported_rabid > rabid_exps, rabid_exps, reported_rabid)
    deaths <- rbinom(length(ttimes), rabid_exps - reported_rabid, prob_death)
    store_deaths[, i] <- deaths
    store_p_rabid[, i] <- p_rabid_t
  }

  results <- cbind(get.meanCI(store_deaths), get.meanCI(store_p_rabid))
  colnames(results) <- c("deaths_mean", "deaths_lowerCI", "deaths_upperCI",
                         "p_rabid_mean", "p_rabid_lowerCI", "p_rabid_upperCI")
  results <- cbind(covars, results)
  return(results)
}

get.burden.dist <- function(covars = mada_district@data, 
                            ttimes = mada_district$ttimes_weighted, 
                            pop = mada_district$pop2015adj, 
                            param_ttimes = B_ttimes_dist,
                            param_intercept = B_0_dist,
                            p_rabid = 0.5, rho_max = 0.9,
                            max_HDR = 25, min_HDR = 5, 
                            dog_rabies_inc = 0.01, human_exp_rate = 0.39, 
                            prob_death = 0.16, nsims = 1000) {

>>>>

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
  pdf_document:
    includes:
      in_header: mystyle.sty
  html_document: default
---

@@

## INTRODUCTION
Deaths due to canine mediated rabies, estimated to cause approximately 60,000 human deaths anually, can be prevented through prompt administration of post-exposure prophylaxis. However, access to this intervention is highly limited in areas where the disease is endemic (*cite GAVI paper or Nandini's paper of geographic availability of PEP*). Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries, with the most commonly available data being numbers of bite victims reporting to health facilities.

@@

The majority of rabies burden studies use these data to estimate burden from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space. While at the national level these estimates may be accurate, at the sub-national level, this framework will likely underestimate rabies deaths in places with low reporting and overestimate rabies deaths in places with high reporting of bites. The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on the level of vaccination coverage and size of the dog population at the national level. Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting.

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we flip the standard decision tree and make the assumption that reported bite incidence reflects access to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing human rabies deaths.

@@

### GIS Data
We used the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs. Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the commune level.

@@

## Centers with no data ## Change this to be so that only ones with zero forms!
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy", "Tsiroanomandidy", 
             "Taolagnaro", "Mandritsara", 
             "Antsiranana", "Marolambo", "Nosy be", "Sainte Marie", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

@@

### Bite patient data
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted as frequently as monthly to annually by clinics. Of the 31 existing ARMC, `r length(no_data) - 2` submitted fewer than 10 forms over the four years. Two clinics, the IPM ARMC and the Fianarantsoa ARMC had separate databases which were not available at the time of analysis. Overall, we had data from `r 31 - length(no_data)` clinics across the country (Fig 1A). These data include the administrative district of the bite patient's address and the date of reporting. We also had `r max(ctar$month, na.rm = TRUE)` months of data from the Moramanga District that were resolved to the commune level (the administrative level below district).

@@

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1). Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC. We excluded any districts from catchments for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).

@@

## To plot lines between ctar + districts
exp_mat %>% 
  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.25
```

```{r figure 1A}
jpeg("figs/network.jpeg")
plot(mada_district, col = mada_district$fill, border = "grey") 
segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, 
         y1 = dist_lines$LATITUDE,
         lwd = dist_lines$size*2, col = dist_lines$color)
points(ctar_points$LONGITUDE, ctar_points$LATITUDE, col = ctar_points$color, pch = 1,
       cex = log(ctar_points$n + 0.1)*0.25)
points(dist_points$dist_long, dist_points$dist_lat, cex = dist_points$size, pch = 20,
       col = dist_points$color)
legend("topright", c("10", "50", "100", "200", "400", "800", "1600"), pch = 1, xjust = 1, yjust = 1,
       pt.cex = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, 
       lwd = log((c(10, 50, 100, 200, 400, 800, 1600)) + 0.1)*0.25, bty = "n", y.intersp = 0.75,
       title = "Number of \n reported bites", xpd = TRUE)
legend("topleft", "A", bty = "n", inset = c(0.2, 0), text.font = 2)
dev.off()
```

```{r network output, echo=FALSE, fig.cap="Figure 1A", results = "show", out.width = '100%'}
knitr::include_graphics("figs/network.pdf")
```

@@

As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2 shows the time series of form submission for each clinic and periods of time excluded by the 15 day threshold). Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold. To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

@@

```{r figure 1B, fig.cap = "Figure 1B"}
clinic_reporting %>%
  mutate(prop15 = prop, prop5 = get.days(threshold = 5)[[1]]$prop,
         prop30 = get.days(threshold = 30)[[1]]$prop) %>%
  ggplot(., aes(x = reorder(ctar, prop), color = ctar)) +
  geom_hline(yintercept = 0.25, color = "darkgrey") +
  geom_boxplot(aes(ymin = prop30, lower = prop15, middle = prop15, upper = prop15, ymax = prop5),
               stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  xlab("") +
  ylab("Proportion of days included") +
  coord_flip() +
  labs(tag = "B")

@@

ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
  scale_color_discrete(name = "Exclude", labels = c("Yes", "No")) +
  ylab("Number of patients per day") +
  geom_jitter(alpha = 0.5, width = 0.25) +
  coord_flip()

plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "n", ylim = c(0, 1), 
     ylab = "Proportion excluded", xlab = "Number of standard deviations")
points(check$sd, check$prop_bites, col = "red", pch = 20)
legend("topright", c("Contacts", "Bites"), col = c("Blue", "Red"), pch = 20, bty = "n")
```

Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure 3B). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)*100, 2)`% of patient records from the national data. We also compared this method to assuming that 20% of total bites at the district level were contacts.

@@

After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).

@@

### Model of reported bites as a function of travel time
While the national bite patient data is available at the district level, travel times and therefore reporting, likely vary significantly at the sub-district level. In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level. Incidence at the commune level is then a function of travel times to the closest ARMC. Specifically, we modeled bites as follows:

@@

Finally, we compared these models to a model of bites at the district level as the function of travel times averaged to the district level:
$$ 
  \mu_{d} = exp(\beta_{t}T_d + \beta_0)\times pop_d 
$$
where $pop_d$, $T_d$, and $mu_d$ are respectively human population size, weighted travel times, and mean number of bites in district $d$. As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models, as a test of whether the observed patterns could be predicted by bite incidence scaling with population size. We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.

@@

### Estimating the impact of expanding PEP Access (TO DO)
We use this framework to compare three scenarios of PEP provisioning in Madagascar:

@@

### Strategically expanding PEP access (TO DO)
Given limited resources and capacity of clinics to provision PEP, we developed a framework to look at the incremental benefit of additional ARMC. Starting with the current locations (Scenario 1), we added one clinic and recalculated the proportion of people living < 3 hours away from a clinic for the country. We added the clinic which minimized this metric, and then repeated the process iteratively, ranking clinics and adding the top clinic in each interation sequentially. We calculate burden for the given clinic locations and look at the incremental reduction in burden as each clinic is added.

### Sensitivity analyses for burden estimates and scenario analyses (TO DO)

@@

par_df <- data.frame(model = c("Mada communes", "Moramanga communes", "Mada districts", 
                               "Mada communes", "Moramanga communes", "Mada districts"),
                     par = c("B_ttimes", "B_ttimes", "B_ttimes", "intercept", "intercept",
                             "intercept"),
                     values = c(B_ttimes_mada, B_ttimes_mora, B_ttimes_dist, 
                                B_0_mada, B_0_mora, B_0_dist),
                     upper_CI = c(CI = B_ttimes_mada_CI[2], B_ttimes_mora_CI[2],
                                  B_ttimes_dist_CI[2],
                                  B_0_mada_CI[2], B_0_mora_CI[2], B_0_dist_CI[2]),
                     lower_CI = c(CI = B_ttimes_mada_CI[1], B_ttimes_mora_CI[1],
                                  B_ttimes_dist_CI[1], 
                                  B_0_mada_CI[1], B_0_mora_CI[1], B_0_dist_CI[1]))
kable(par_df, caption = "Table 1: summary of parameter estimates")
ttimes_plot <- seq(0, 15*60, by = 30)
pop_plot <- 1e5

@@

legend("topright", c("Mada communes", "Mada districts", "Moramanga communes"), 
       col = c("darkblue", "purple", "red"), 
       lty = c(1, 1, 1), 
       bty = "n")
```

@@

```{r figure 5B, fig.cap="Figure 5B"}
## Data vs. predicted
comm_data %>%
      filter(exclude == 0) %>%
      mutate(bites = exp(B_ttimes_mada*ttimes_weighted + B_0_mada)*pop) %>%
      group_by(mdg_dis_co) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(exps_dist, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites, bty = "n", pch = 20, col = alpha("darkblue", 0.5),
                   xlab = "Observed mean bites", ylab = "Model predicted mean bites", 
                   xlim = c(0, 5000))
abline(a = 0, b = 1, col = "grey", lty = 2)

exps_dist %>%
      mutate(bites_mu = exp(B_ttimes_dist*ttimes_weighted_dist + B_0_dist)*pop) -> check
points(check$bites, check$bites_mu, col = alpha("purple", 0.5), pch = 20)

mora_bites %>%
      mutate(bites_mu = exp(B_ttimes_mora*ttimes_weighted + B_0_mora)*pop) -> check
points(check$bites, check$bites_mu, bty = "n", pch = 20, col = alpha("red", 0.5))
legend("topright", inset = c(-0.1, -0.2),
       c("Mada communes", "Mada districts", "Moramanga communes"), 
       col = c(alpha("darkblue", 0.5), alpha("purple", 0.5), alpha("red", 0.5)), 
       pch = 20, 
       bty = "n", xpd = TRUE)

@@

We estimated similar parameter values from our commune-level data from the Moramanga ARMC and the district level data from 19 clinics across the country (Table 1, Figure 5A), with reported bite incidence decreasing with travel times to the ARMC. All of the models produced reasonable fits to the data (Figure 5B), however, there was some variation in bite incidence that was not captured by the model.

#### Model validation: TO DO

@@

```{r burden results, fig.cap = "Figure 7", cache = TRUE}
## Commune level
source("R/get.burden.R")
results_pr20 <- get.burden(p_rabid = 0.20, rho_max = 0.98)
results_pr60 <- get.burden(p_rabid = 0.60,rho_max = 0.98)
results_pr20$pr <- "0.20"
results_pr60$pr <- "0.60"
results <- rbind(results_pr20, results_pr60)
